---
title: "Final Project"
author: "Peter Siw"
date: "29/05/2018"
output: ioslides_presentation
transition: "rotate"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(reshape2)
library(grid)
aq <- airquality[complete.cases(airquality), ]
```

## Model Overfit

My App (https://petersiw.shinyapps.io/final_project/) discusses the 
problem of overfitting a multivariate model due to inclusion of too many 
variables by comparing between the R Squared ($R^2$) and Adjusted R Squared ($\bar{R^2}$).

Using the Airquality data, I created a simple model analysing the relationship
between Ozone Levels and Temperature. By adjusting the slider, the user can
simulate a model with varying numbers of variables (which are simply the 
exponents of Temperature). 

The model used was $Ozone~=~\beta_{0}~+~\beta_{1}Temp~+~\beta_{k}Temp^k~+~...~+\beta_{n}Temp^n$ where $n$ is being varied.

## Plots of the Model Across Different $n$ Degrees of Polynomial

```{r, echo=FALSE}
#polynomial degree = 3
p1 <- ggplot(data = aq, aes(x = Temp, y = Ozone)) + 
      geom_point(colour = "steelblue") + 
      labs(title = "n = 3",
           x = "Temperature", y = "Ozone Level") + 
      theme(plot.title = element_text(hjust = 0.5)) +
      geom_smooth(method = "lm", 
                  formula = y ~ poly(x, 3),
                  colour = "red", se = F)
#polynomial degree = 7
p2 <- ggplot(data = aq, aes(x = Temp, y = Ozone)) + 
      geom_point(colour = "steelblue") + 
      labs(title = "n = 7",
           x = "Temperature", y = "Ozone Level") + 
      theme(plot.title = element_text(hjust = 0.5)) +
      geom_smooth(method = "lm", 
                  formula = y ~ poly(x, 7),
                  colour = "red", se = F)
#create r squared and adj r squared plot
n <- seq(1, 15, by = 1)
df1 <- data.frame(n = c(1:length(n)), r2 = rep(NA, length(n)), 
                  adj_r2 = rep(NA, length(n)))
colnames(df1) <- c("n", "R Squared", "Adjusted R Squared")
for(i in 1:length(n)) {
  m <- lm(formula = (formula = Ozone ~ poly(Temp, i)), data = aq)
  df1[i, 2] <- summary(m)$r.squared
  df1[i, 3] <- summary(m)$adj.r.squared
}
df2 <- melt(df1, id = "n")
p3 <- ggplot(data = df2, aes(x = n, y = value, colour = variable)) +
      geom_line() +
      labs(title = "Plot of R Squared and Adjusted R Squared against the 
           Number of Variables", x = "Number of Variables", 
           y = "R Squared and Adjusted R Squared") + 
      theme(plot.title = element_text(hjust = 0.5)) +
      theme(legend.position = "bottom", legend.title = element_blank()) +
      geom_vline(xintercept = 4, colour = "orchid4", linetype = "twodash")
pushViewport(viewport(layout = grid.layout(nrow = 2, ncol = 2)))
view <- function(x, y) viewport(layout.pos.row = x, layout.pos.col = y)
print(p1, vp = view(1, 1))
print(p2, vp = view(2, 1))
print(p3, vp = view(1:2, 2))
```

## An Example of Code for the Previous Slide

``` {r, echo = T}
p4 <- ggplot(data = aq, aes(x = Temp, y = Ozone)) + 
      geom_point(colour = "steelblue") + 
      labs(title = "n = 7", x = "Temperature", y = "Ozone Level") + 
      theme(plot.title = element_text(hjust = 0.5)) +
      geom_smooth(method = "lm", formula = y ~ poly(x, 7),
                  colour = "red", se = F)
```

## Analysis and Conclusion

From the first plot, we can see that once we arrive at $n~=~5$, the model does not really change. It just increase slightly in complexity without much fundamental shape changes, unlike that between $n~=~1$ and $n~=~3$. 

As seen from the plot above, $\bar{R^2}$ decreases rapidly after $n~=~4$. This shows that any model bigger than that does not explain more of the population without introducing more bias to the model.

This is a classic case of overfitting, whereby the model is becoming too close to the data that it becomes biased. Therefore, one must always monitor the $\bar{R^2}$ of a model to prevent overfitting.
